---
title: "Linh tinh"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2025-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyquant)

dataset <- tq_get('AAPL',
               from = "2023-01-01",
               to = "2025-01-01",
               get = "stock.prices")

dataset <- dataset[, c("date", "adjusted")]
#dataset$x <- dataset$adjusted
#dataset
dataset$x <- log(dataset$adjusted)
head(dataset)


#dataset <- read.csv("C:\\Users\\khuem\\Downloads\\arimadat(1).csv")
#dataset
```
                            series x is the log transform of the adjusted close price.

```{r}
n <- length(dataset$x)
data_train <- dataset$x[1:(n - 25)] 
data_test <- dataset$x[(n - 25 + 1):n] 
plot(data_train, type = 'l')
acf(data_train)
pacf(data_train)
```
                            From the Series Plot and the PACF and ACF, the series is highly non-stationary. Therefore, taking difference is needed.


```{r}
diff1 <- diff(data_train)
diff1[1:10]
plot(diff1, type = 'l') 
```
                                 Diff1 series look much more stationary than the original data. However, there is still a visible trend in the data. Therefore, taking the 2nd order difference is needed.

```{r}

diff2 <- diff(diff1)
plot(diff2, type = 'l') 
pacf(diff2)

ar_demo <- ar(diff2, aic = TRUE, method = 'mle')
ar_demo
```
                                  The diff2 plot looks very stationary. 
                                  
                                  PACF plot of the diff2 series suggests that arima order 11 or 12 should be good.
                                  
                                  Fitting ar function model to determine the order for the ARIMA model. The order selected for the arima model by the ar function is 12. Therefore, later a complete ARIMA(12,2,0) will be fitted on the original adjusted close price for forecasting purpose.

```{r}
ar_mle <- arima(data_train, order = c(12,2,0))
ar_mle

resids <- ar_mle$resid
plot(resids, type = 'l')
pacf(resids)

library(tseries)
adf_test <- adf.test(resids, alternative = "stationary")
print("ADF Test:")
print(adf_test)
```
                                  The model above is the fitted ARIMA(12,2,0) model.
                                  
                                  The residuals of the ARIMA model strongly resemblances a white noise process, thus proving that the ARIMA(12,2,0) is enough to capture all the trend of the data series.
                                  
                                  The PACF of the residuals also proves the above statement. 
                                  
                                  To be more confirmative, Augmented Dickey-Fuller Test is conducted. The null hypothesis of non-stationary series is rejected with a p-value = 0.01 (highly significant). The conclusion is stationary.

```{r}
library(forecast);
fitted_values <- fitted(ar_mle)
fitted_values[1:10]
```
```{r}

mypreds <- predict(ar_mle, n.ahead = 25, se.fit = TRUE)
n <- length(data_train)
fore <- c(data_train[(n-90):n], mypreds$pred)
foreupper <- c(data_train[(n-90):n], mypreds$pred + 2 * mypreds$se)
forelower <- c(data_train[(n-90):n], mypreds$pred - 2 * mypreds$se)
plot(fore, type = 'l',
     ylim = range(c(foreupper, forelower, data_test)), 
     main = "Prediction vs Actual Test Data (Last 90 Points)", ylab = "Values")
lines(foreupper, lty = 2, col = 'red')
lines(forelower, lty = 2, col = 'red')
lines((length(data_train[(n-90):n]) + 1):(length(data_train[(n-90):n]) + length(data_test)), 
      data_test, col = 'blue', type = 'l', lwd = 2)


fore <- c(data_train, mypreds$pred)
foreupper <- c(data_train, mypreds$pred + 2 * mypreds$se)
forelower <- c(data_train, mypreds$pred - 2 * mypreds$se)
plot(fore, type = 'l',
     ylim = range(c(foreupper, forelower, data_test)),
     main = "Prediction vs Actual Test Data (Full Series)",
     ylab = "Values", xlab = "Time")
lines(foreupper, lty = 2, col = 'red')
lines(forelower, lty = 2, col = 'red')
lines((n + 1):(n + length(data_test)),
      data_test, col = 'blue', type = 'l', lwd = 2)




```

```{r}
# performance metrics for diff2 arima model
print('Performance metrics for diff2 arima')
mae <- mean(abs(data_test - mypreds$pred))  
mse <- mean((data_test - mypreds$pred)^2)  
rmse <- sqrt(mse)  
mape <- mean(abs((data_test - mypreds$pred) / data_test)) * 100  
ss_res <- sum((data_test - mypreds$pred)^2)  
ss_tot <- sum((data_test - mean(data_test))^2)  
r_squared <- 1 - (ss_res / ss_tot) 

cat("MAE:", mae, "\n")
# cat("MSE:", mse, "\n")
cat("RMSE:", rmse, "\n")
cat("MAPE:", mape, "%\n")
cat("R-squared:", r_squared, "\n")
```
                              The error metrics (MAE, MSE, RMSE, MAPE) suggest that the model’s predictions are relatively accurate.





